Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-12 paper/paper.tex
Changes from Sonja in Overleaf

\paragraph{Graphs.} Let $G$ and $H$ be graphs. Let $v \in \vertexset(G)$ be a vertex of $G$. The
\emph{(open) neighbourhood} of $v$, written $\neighbourhood_G(v)$, is the set of vertices adjacent
to $v$ not including $v$ itself, whilst the closed neighbourhood of $v$, written
$\neighbourhood_G[v]$, is the neighbourhood plus $v$. The \emph{degree} of a vertex, $\degree_G(v)$,
is the cardinality of its open neighbourhood. The \emph{neighbourhood degree sequence} of a vertex
is the sequence consisting of the degrees of its neighbours, in descending order. Given a vertex set
$S \subseteq \vertexset(G)$, the subgraph \emph{induced by} $S$, written $G[S]$, is the subgraph of
$G$ with only the vertices in $S$ together with all the edges between them. A \emph{clique} is a
subgraph where every vertex is adjacent to every other in the subgraph.

@@

\paragraph{Degree-based invariants for homomorphisms.} The use of constraint programming techniques
for finding homomorphisms relies upon exploiting various invariants. For example, every recent
constraint programming solver makes use of the following result.

\begin{proposition}\label{proposition:degreends}
    Suppose a subgraph isomorphism $i$ exists from a graph $G$ to a graph $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item The function cannot map $v$ to a vertex of lower degree, $\degree(v) \le \degree(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own
            \cite{DBLP:journals/constraints/ZampelliDS10}.
    \end{enumerate}
\end{proposition}

@@

These invariants do \emph{not} hold for homomorphisms in general. As a simple counter-example, a
homomorphism may map a star graph (one central vertex of high degree, which is adjacent to many
other vertices, each of which has no other neighbours) onto a single edge. However, it is
straightforward to verify the following.

\begin{proposition}Both invariants in \cref{proposition:degreends} are also valid for locally
injective homomorphisms.\end{proposition}

@@

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at most the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

@@

\begin{proposition}\label{proposition:clique}
    Suppose we are finding a graph homomorphism $i$ from a graph $G$ to a graph $H$ where $H$ does
    not contain any loops. Suppose there is a $k$ vertex clique in $G$. Then the image of this
    clique under $i$ must also be a $k$ vertex clique.
\end{proposition}

>>>>

\paragraph{Graphs.} Let $G$ and $H$ be graphs. Let $v \in \vertexset(G)$ be a vertex of $G$. The
\emph{(open) neighbourhood} of $v$, written $\neighbourhood_G(v)$, is the set of vertices adjacent
to $v$ not including $v$ itself, whilst the closed neighbourhood of $v$, written
$\neighbourhood_G[v]$, is the neighbourhood of $v$ plus $v$. The \emph{degree} of a vertex, $\degree_G(v)$,
is the cardinality of its open neighbourhood. The \emph{neighbourhood degree sequence} of a vertex
is the sequence consisting of the degrees of its neighbours, in descending order. Given a vertex set
$S \subseteq \vertexset(G)$, the subgraph \emph{induced by} $S$, written $G[S]$, is the subgraph of
$G$ with only the vertices in $S$ together with all the edges between them. A \emph{clique} is a
subgraph where every vertex is adjacent to every other in the subgraph.

@@

\paragraph{Degree-based invariants for homomorphisms.} The use of constraint programming techniques
for finding homomorphisms relies upon exploiting various invariants. It is straightforward to verify the following.

\begin{proposition}\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from a $G$ to  $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own
           .
    \end{enumerate}
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these also hold for subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}.
Every recent
constraint programming solver makes use of the of the above invariants for subgraph isomowphisms.
However, these invariants do \emph{not} hold for homomorphisms in general. As a simple counter-example, a
homomorphism may map a star graph (one central vertex of high degree, which is adjacent to many
other vertices, each of which has no other neighbours) onto a single edge.

@@

@@

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at least the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

@@

\begin{proposition}\label{proposition:clique}
    Let $i$ be a homomorphism from $G$ to $H$ where $H$ does
    not contain any loops. Let $S\subset V$ be such that $G[S]=K_k$
\end{proposition}
